---
title: "Practice KNN A3.202-v1"
author: Dr. Nishat Mohammad
date: 01/24/2024
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

## Task 1: Loading Prostate Cancer Dataset
Download the data set for the tutorial and save it in your project folder.  

```{r}
# Load data to variable
prsca_data1 <- (read.csv("Prostate_Cancer.csv", stringsAsFactors = FALSE))

# Look at the data
str(prsca_data1)

# Check dimensions
dim(prsca_data1)

# Check for missing values
any(is.na(prsca_data1))
# FALSE means there are no missing values!

```
## Task 2:
Follow this tutorial on applying kNN to prostate cancer detection and implement all of the steps in your R Notebook. Use appropriate headers to each step to structure your notebook. Make sure to explain each step and what it does. (Note: The data set provided as part of this assignment has been slightly modified from the one used in the tutorial, so small deviations in the result can be expected.).  

### Take off the id column
```{r ID_Off}
# Take the id  column (first column) out of the data frame
prsca_data1 <- prsca_data1[-1]

# Look at the editted data 
str(prsca_data1)

```
The ID column has been removed leaving us with 9 variables.

### Factor for Beingn and Malignant Tumor Catogories
```{r}
# Factoring step
prsca_data1$diagnoses <- factor(prsca_data1$diagnosis_result, levels = c("B", "M"), labels = c("Benign", "Malignant"))
#str(prsca_data1)

# Get the percentages
round(prop.table(table(prsca_data1$diagnosis)) * 100, digits = 1)

```
This shows that Benign tumors contribute to 38% of the observations while malignant contribute to 68% of the observations.  

### Normalizing numeric data
From the structure we can see the  that 1st to 4th columns are integers and 5th to 8th columns are numeric.  
```{r Summary of data}
# Check the summary of the data except the first and last columns
summary(prsca_data1[2:9])
```
This give me a quick look at all the statistics.  

```{r Normalization}
# Make a function
nrm <- function(x) {
return ((x - min(x)) / (max(x) - min(x)))
}

# Normalize
prsca_data_nrm <- as.data.frame(lapply(prsca_data1[2:9], nrm))
str(prsca_data_nrm)
summary(prsca_data_nrm$perimeter)
```
Now all the columns are numeric and values normalized. I checked the perimeter summary, values are normalized.   

###  Creating Training and Test Data Set
This will be done with ratio of 65 for training and 35 for test data.  

```{r test_train_sets}
# divide the data set into 2 portions
# in the ratio of 65: 35 (assumed) 
# for the training and test data set respectively
training_set <- prsca_data_nrm[1:65,]
test_set <- prsca_data_nrm[66:100,]

```
  
  
Now I will add labels for the training and test sets
```{r add_labels}
# Make the labels in accordance with the diagnosis_result column (column 1)

training_labels <- prsca_data1[1:65, 1]
testing_labels <- prsca_data1[66:100, 1]
```

### Training a model on data
```{r}
library(class)
# Get value of k
#nrow(prsca_data1)
K <- sqrt(nrow(prsca_data1))
#any(is.na(prsca_data1))
# apply knn() function
test_knn_prediction <- knn(train = training_set, test = test_set, cl = training_labels, k= K)


```
Here, I found the k by taking the square root of the total number of observations. k is 10 form the 100 observations in this dataset.  
I applied the knn function here using the both sets and the labels of the training set with k of 10. Now let us evaluate the model.  


### Evaluate the model performance
```{r}
library(gmodels)

# use the cross table
prsc_cros_table <- CrossTable(
  x = testing_labels,
  y = test_knn_prediction,
  prop.chisq=FALSE)
```
The Cross table above shows the prediction of cancer samples in the data set provided.  
There are a total of 19 benign observations in the test set, out of these our model predicted 7 correctly (true positive) and 12 incorrectly as malignant (false negative).    
There are total of 16 malignant observations in the test set, out of there our model predicted 15 correctly (true negative) and 1 incorrectly as benign (false positive).  

The overall accuracy of the model on the test set is the percentage of sum of true events fraction which is 7+ 15 divide by 35 which is  approximately 62.9%. This is the precision of the model.  


### Fluctuate the value of K around 10 to check for better accuracy:  

```{r k11model}
K11 <- 11
test_knn_prediction_k11 <- knn(train = training_set, test = test_set, cl = training_labels, k= K11)
prsc_cros_table_k11 <- CrossTable(
  x = testing_labels,
  y = test_knn_prediction_k11,
  prop.chisq=FALSE)
```

From this cross table we can see the true event fraction reduce to (5+16)/35  which is 60.0% this is the accuracy or precision of the model

```{r k9model}
K9 <- 9
test_knn_prediction_k9 <- knn(train = training_set, test = test_set, cl = training_labels, k= K9)
prsc_cros_table_k9 <- CrossTable(
  x = testing_labels,
  y = test_knn_prediction_k9,
  prop.chisq=FALSE)
```

For this model we can see that the true event fraction is increased to (8 + 16)/35  which is 68.6% accuracy.

My conclusion is that the knn model will be more accurate if k is reduced. but the value of k should be carefully decided in order not to leave out some trivial patterns which may be equally important for later interpretation.

## Task 3:
Once you’ve complete the tutorial, try another kNN implementation from another package, such as the caret package. Compare the accuracy of the two implementations.  

```{r KNN_with_caret }
# Load data 
p_df <- read.csv("Prostate_Cancer.csv",stringsAsFactors=FALSE)
str(p_df)
# Take off id column
p_df <- p_df[-1]

# Load caret package
library(caret)

# For reproducibility
set.seed(786)

# Train and test sets
caret_training_set <- createDataPartition(y = p_df$diagnosis_result, p= 0.64, list = FALSE)
train_df <- p_df[caret_training_set,]

testing_df <- p_df[-caret_training_set,]

# Factor the data according to tumor categories benign and malignant
train_df[["diagnosis_result"]] = factor(train_df[["diagnosis_result"]])
testing_df[["diagnosis_result"]] = factor(testing_df[["diagnosis_result"]])


# Normalize and Train the model
#?trainControl
train_control <- trainControl(method = "repeatedcv", number=10, repeats = 3)

knn_pred_model <- train(diagnosis_result ~., data = train_df, method = "knn",
                 trControl=train_control,
                 preProcess = c("center", "scale"),
                 tuneLength = 20)
knn_pred_model
# Visualize the prediction
plot(knn_pred_model)

# Use the model on test set
knn_test_pred <- predict(knn_pred_model, newdata = testing_df)

```

## Task 4:
Use the confusionMatrix() function from the caret package to determine the accuracy of both algorithms.  

```{r confusion_matrix}

# test  the knn from caret package
length(testing_df$diagnosis_result)
length(knn_test_pred)

confusionMatrix(table(knn_test_pred, testing_df$diagnosis_result))
```
The accuracy of the KNN model using caret is 0.8571 when considering the benign Class is the positive class. 

```{r confusion_matrix_class}
confusionMatrix(table(knn_test_pred, testing_labels))
```
The accuracy of this model is 0.28.  
Comparing the accuracy of both models, the knn by the caret package has a higher accuracy.

















